---
title: "Decision Support Tool - Mixed Models"
format:
  pdf:
      fontsize: 8pt
---



I cobbled together the models from Elena's recommendations.

The main feedback on the statistics was to treat the study arms as a 2x2 factorial design:
```
| Baseline           | Prediction support |
|--------------------|--------------------|
| Case-based support | Combined support   |
```

packages:

- lme4
- DHARMa
- emmeans
- ggplot2
- pbkrtest
- lmerTest

```{r}
#| message: false
#| warning: false
#| code-fold: true
#| code-summary: "Libraries, some functions"
library(lme4)
library(emmeans)
library(DHARMa)
library(ggplot2)


# A function to put results in a table
glmer_markdown <- function(model, caption = "") {
    s <- summary(model)
    coefs <- as.data.frame(s$coefficients)

    OR <- exp(coefs$Estimate)
    CI_lower <- exp(coefs$Estimate - 1.96 * coefs$`Std. Error`)
    CI_upper <- exp(coefs$Estimate + 1.96 * coefs$`Std. Error`)

    res_df <- data.frame(
        Term = rownames(coefs),
        Estimate = sprintf("%.3f", coefs$Estimate),
        `Std. Error` = sprintf("%.3f", coefs$`Std. Error`),
        `z value` = sprintf("%.3f", coefs$`z value`),
        `p value` = ifelse(coefs$`Pr(>|z|)` < 0.001, "< 0.001", sprintf("%.3f", coefs$`Pr(>|z|)`)),
        `OR (95% CI)` = sprintf("%.2f [%.2f, %.2f]", OR, CI_lower, CI_upper),
        check.names = FALSE
    )

    knitr::kable(res_df, caption = caption, align = "lrrrrr")
}

lmer_markdown <- function(model, caption = "") {
    s <- summary(model)
    coefs <- as.data.frame(s$coefficients)

    CI_lower <- coefs$Estimate - 1.96 * coefs$`Std. Error`
    CI_upper <- coefs$Estimate + 1.96 * coefs$`Std. Error`
    OR <- exp(coefs$Estimate)
    res_df <- data.frame(
        Term = rownames(coefs),
        Estimate = sprintf("%.4f", coefs$Estimate),
        `Std. Error` = sprintf("%.4f", coefs$`Std. Error`),
        `95% CI` = sprintf("[%.4f, %.4f]", CI_lower, CI_upper),
        `t value` = sprintf("%.3f", coefs$`t value`),
        check.names = FALSE
    )

    if ("Pr(>|t|)" %in% colnames(coefs)) {
        res_df$`p value` <- ifelse(coefs$`Pr(>|t|)` < 0.001, "< 0.001", sprintf("%.4f", coefs$`Pr(>|t|)`))
    } else {
        p_vals <- 2 * (1 - pnorm(abs(coefs$`t value`)))
        res_df$`p value)` <- ifelse(p_vals < 0.001, "< 0.001", sprintf("%.4f", p_vals))
    }

    res_df$odds_ratio <- OR <- round(exp(coefs$Estimate), 3)

    knitr::kable(res_df, caption = caption, align = "lrrrrr")
}
```





## Logistic mixed model for exact match

I'm not qualified to say if the exact diagnostic match is clinically meaningful.
But here is a logistic mixed model for this primary outcome.

> **Feedback**: Discuss the jaccard model but note problems with the data and Goodness of Fit. Focus on the exact diagnostic match. Discuss sensitivity analyses but non of them were significant.

### Preprocessing
```{r}
#| code-fold: true
#| code-summary: "Dataframe preprocessing (adding the factors etc)"
raw_data <- read.csv("records.csv")

# Check if prediction is correct for every row
raw_data$correct <- raw_data$pred == raw_data$gold

# Group data to patient-case level (540 evaluations: 27 participants x 20 patients)
# Calculate trial_order (1 to 20) per participant based on unique patient appearance sequence
df_cases_list <- list()

participants <- unique(raw_data$participant)

for (p in participants) {
    p_data <- raw_data[raw_data$participant == p, ]
    # unique patient sequence for participant
    patient_seq <- unique(p_data$patient)

    for (t in seq_along(patient_seq)) {
        pat_id <- patient_seq[t]
        sub <- p_data[p_data$patient == pat_id, ]

        fully_corr <- all(sub$correct)
        jaccard_val <- sub$jaccard[1]
        arm_val <- sub$arm[1]
        exp_level <- sub$experience_level[1]

        df_cases_list[[length(df_cases_list) + 1]] <- data.frame(
            participant = as.factor(p),
            patient = as.factor(pat_id),
            arm = arm_val,
            trial_order = t,
            experience_level = as.factor(exp_level),
            fully_correct = fully_corr,
            jaccard = jaccard_val,
            stringsAsFactors = FALSE
        )
    }
}


df_cases <- do.call(rbind, df_cases_list)
# Create 2x2 Factorial factors
df_cases$pred_support <- factor(ifelse(df_cases$arm %in% c(1, 3), 1, 0), labels = c("No", "Yes"))
df_cases$case_support <- factor(ifelse(df_cases$arm %in% c(2, 3), 1, 0), labels = c("No", "Yes"))
df_cases$arm_factor <- factor(df_cases$arm,
    levels = c(0, 1, 2, 3),
    labels = c("Baseline", "PredSupport", "CaseSupport", "Combined")
)

# Add multi-label factor
patient_gold_sums <- aggregate(gold ~ patient, data = raw_data, FUN = function(x) sum(x) / (length(x) / 7))
multilabel_patients <- patient_gold_sums$patient[patient_gold_sums$gold > 1]
df_cases$is_multilabel <- factor(ifelse(df_cases$patient %in% multilabel_patients, "Multi-label", "Single-label"))


# Calculate case difficulty (for one of the sensitivity analyses) from the baseline arm
baseline_cases <- subset(df_cases, arm == 0)
case_diff_df <- aggregate(fully_correct ~ patient, data = baseline_cases, FUN = mean)
colnames(case_diff_df)[2] <- "baseline_case_accuracy"
df_cases <- merge(df_cases, case_diff_df, by = "patient", all.x = TRUE)

df_display <- data.table::copy(head(df_cases, 10))
df_display$participant = "<blanked>"
if (knitr::is_html_output()) {
    rmarkdown::paged_table(df_display)
}
```



``` {r}
cat("Number of evaluations:", nrow(df_cases))
cat("Total participants:", length(unique(df_cases$participant)))
```


### Main model: 2x2 exact diagnostic match
Per Elena's feedback:

>_Ideally, fit
an appropriate mixed-effects model with participant and ECG case as crossed random effects. If
exact diagnostic match is clinically meaningful, use a logistic mixed model for that primary
outcome and retain Jaccard as a secondary measure. Include trial order and expertise only where
justified._
```{r}
m1_primary <- glmer(
    fully_correct ~ pred_support * case_support + (1 | participant) + (1 | patient),
    data = df_cases,
    family = binomial
)

glmer_markdown(m1_primary, caption = "Main Model (Exact match)")
cat("\nRandom Effects Variance:\n")
VarCorr(m1_primary)
```

**Is the model fitted properly?**
We can check this with a simulated residuals plot with the `DHARMa` library:
```{r}
#| fig-width: 12
#| fig-height: 6
#| out-width: "100%"
#| fig-align: "center"
sim_res <- simulateResiduals(fittedModel = m1_primary, n = 250)
plot(sim_res)
```

### Model 2: GLLM adjusted for `trial_order` and `experience` covariates

```{r}
m2_covariates <- glmer(
    fully_correct ~ pred_support * case_support + trial_order + experience_level + (1 | participant) + (1 | patient),
    data = df_cases,
    family = binomial
)
glmer_markdown(m2_covariates, caption = "Adjusted for trial_order and experience: Fixed Effects & Odds Ratios")
```

**Is the model fitted properly?**
```{r}
#| fig-width: 12
#| fig-height: 6
#| out-width: "100%"
#| fig-align: "center"
sim_res <- simulateResiduals(fittedModel = m2_covariates, n = 250)
plot(sim_res)
```

### Sensitivity analysis 1: exclude participants without interval

Two participants (P1 and P2) did not have the interval measurement feature.
It was possible to approximate the distance from the plot grid, however.

```{r}
m_sens1 <- glmer(
    fully_correct ~ pred_support * case_support + (1 | participant) + (1 | patient),
    data = subset(df_cases, !participant %in% c("5", "7")),
    family = binomial
)
glmer_markdown(m_sens1, caption = "Sensitivity analysis 1: Exclude P1 and P2")

cat("\nRandom Effects Variance:\n")
VarCorr(m_sens1)
```

**Is the model fitted properly?**
```{r}
#| fig-width: 12
#| fig-height: 6
#| out-width: "100%"
#| fig-align: "center"
sim_res <- simulateResiduals(fittedModel = m_sens1, n = 250)
plot(sim_res)
```

### Sensitivity analysis 2: Single vs Multi-label

```{r}
single_sub <- subset(df_cases, is_multilabel == "Single-label")
multi_sub <- subset(df_cases, is_multilabel == "Multi-label")
cat("Single label n =", nrow(single_sub), "\nMulti label n =", nrow(multi_sub), "\n")
```
Error is much larger on the multi-label subset, but its number of samples is twice as small.

#### 2A: Single-label
```{r}
m_sens2_single <- glmer(
    fully_correct ~ pred_support * case_support + (1 | participant) + (1 | patient),
    data = subset(df_cases, is_multilabel == "Single-label"),
    family = binomial
)
glmer_markdown(m_sens2_single, caption = "Sensitivity 2a: Single-label Cases Subgroup")
```

```{r}
#| fig-width: 12
#| fig-height: 6
#| out-width: "100%"
#| fig-align: "center"
sim_res <- simulateResiduals(fittedModel = m_sens2_single, n = 250)
plot(sim_res)
```

#### 2B: Multi-label
```{r}

m_sens2_multi <- glmer(
    fully_correct ~ pred_support * case_support + (1 | participant) + (1 | patient),
    data = subset(df_cases, is_multilabel == "Multi-label"),
    family = binomial
)

glmer_markdown(m_sens2_multi, caption = "Sensitivity 2b: Multi-label Cases Subgroup")
```

```{r}
#| fig-width: 12
#| fig-height: 6
#| out-width: "100%"
#| fig-align: "center"
sim_res <- simulateResiduals(fittedModel = m_sens2_multi, n = 250)
plot(sim_res)
```

### Sensitivity analysis 3: Case difficulty & trial order

``` {r}
m_sens3 <- glmer(
    fully_correct ~ pred_support * case_support + trial_order + baseline_case_accuracy + (1 | participant) + (1 | patient),
    data = df_cases,
    family = binomial
)

glmer_markdown(m_sens3, caption = "Sensitivity 3: case difficulty and trial order")
```

```{r}
#| fig-width: 12
#| fig-height: 6
#| out-width: "100%"
#| fig-align: "center"
sim_res <- simulateResiduals(fittedModel = m_sens3, n = 250)
plot(sim_res)
```

### Sensitivity analysis 4: Exclude patient 655

Results of patient 655 seemed a bit off.
Olof recommended not mentioning it.

```{r}
m_sens4 <- glmer(
    fully_correct ~ pred_support * case_support + (1 | participant) + (1 | patient),
    data = subset(df_cases, patient != "655"),
    family = binomial
)
glmer_markdown(m_sens4, caption = "Sensitivity 4: exclude patient 655")
```

```{r}
#| fig-width: 12
#| fig-height: 6
#| out-width: "100%"
#| fig-align: "center"
sim_res <- simulateResiduals(fittedModel = m_sens4, n = 250)
plot(sim_res)
```

## LMM on Jaccard

### Model 3: Linear Mixed Model on Jaccard

```{r}
m3_jaccard <- lmer(
    jaccard ~ pred_support * case_support + (1 | participant) + (1 | patient),
    data = df_cases
)

lmer_markdown(m3_jaccard, caption = "Model 3: Linear Mixed Model on Jaccard Index")
```

**Is the model fitted properly?** This doesn't seem to be the case
```{r}
#| fig-width: 12
#| fig-height: 6
#| out-width: "100%"
#| fig-align: "center"
sim_res <- simulateResiduals(fittedModel = m3_jaccard, n = 250)
plot(sim_res)
```


## All results in order

```{r}
glmer_markdown(m1_primary, caption = "Main Model (Exact match)")
glmer_markdown(m2_covariates, caption = "Adjusted for trial_order and experience")
glmer_markdown(m_sens1, caption = "Sensitivity 1: Exclude P1 and P2 (no interval feature)")
glmer_markdown(m_sens2_single, caption = "Sensitivity 2a: Single-label Cases Subgroup")
glmer_markdown(m_sens2_multi, caption = "Sensitivity 2b: Multi-label Cases Subgroup")
glmer_markdown(m_sens3, caption = "Sensitivity 3: case difficulty and trial order")
glmer_markdown(m_sens4, caption = "Sensitivity 4: exclude patient 655")
```

### Pairwise comparisons between study arms

#### Main model (Exact match)
``` {r}
model_arms <- glmer(
    fully_correct ~ arm_factor + (1 | participant) + (1 | patient),
    data = df_cases,
    family = binomial
)
emm <- emmeans(model_arms, ~arm_factor, type = "response")
print(pairs(emm))
```

#### Jaccard model
``` {r}
model_arms <- lmer(
    jaccard ~ arm_factor + (1 | participant) + (1 | patient),
    data = df_cases,
)
emm <- emmeans(model_arms, ~arm_factor, type = "response")
print(pairs(emm))
```

```{r}
#| fig-width: 12
#| fig-height: 6
#| out-width: "100%"
#| fig-align: "center"
sim_res <- simulateResiduals(fittedModel = model_arms, n = 250)
plot(sim_res)
```