---
title: "Qualitative Thematic Analysis"
subtitle: "Post-study survey — CHI 2027 Florentine"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    code-tools: true
    code-overflow: wrap
    embed-resources: true
    theme: cosmo
execute:
  echo: true
  warning: false
  message: false
  error: false
editor: source
---

<style>
body { line-height: 1.55; }
.callout { margin-top: 1.2rem; margin-bottom: 1.2rem; }
table { font-size: 0.90rem; }
blockquote { border-left-color: #2c5282; color: #243447; }
.cell-output-display { overflow-x: auto; }
</style>

## Purpose and analytic status

This executable notebook contains **only the qualitative analysis** of the two
open-ended post-study survey questions:

1. *What was your experience with the tool?*
2. *What factors made you trust or distrust the AI's prediction?*

It implements a semantic codebook thematic analysis of the updated survey with
27 participants. Twenty-six participants supplied at least one open-ended
response; one participant completed the closed-ended survey but supplied no
qualitative text. The notebook does not analyze closed-ended survey items,
diagnostic performance, or experimental-condition effects.

To render the notebook in RStudio, open this file and click **Render**. From a
terminal, run:

```bash
quarto render thematic_analysis_notebook.qmd
```

The result is a self-contained HTML file in which every R code block is shown
immediately above its output. The notebook requires R 4.1 or newer, Quarto, and
the `knitr` package normally installed with RStudio/Quarto support; the analysis
itself uses no additional R packages.

## Configuration and reproducibility safeguards

The exact OneDrive directory supplied for the project is the default. The
environment variable `CHI2027_FLORENTINE_DIR` can override it when the project
is run on another computer. If the default directory does not exist but the CSV
is next to the notebook, the notebook uses its current directory.

```{r}
#| label: configuration

DEFAULT_PROJECT_DIR <- paste0(
  "/Users/elenaberetta/Library/CloudStorage/",
  "OneDrive-VrijeUniversiteitAmsterdam/",
  "Documenti - OneDrive/VU/Codice/CHI2027-Florentine"
)

project_dir <- Sys.getenv(
  "CHI2027_FLORENTINE_DIR",
  unset = DEFAULT_PROJECT_DIR
)

if (!dir.exists(project_dir)) {
  local_candidates <- list.files(
    getwd(),
    pattern = "^survey_data_all_participants.*\\.csv$",
    full.names = TRUE,
    ignore.case = TRUE
  )

  if (length(local_candidates) == 0L) {
    stop(
      paste0(
        "Project directory not found:\n", project_dir,
        "\n\nPlace the notebook and survey CSV in the same directory, ",
        "edit DEFAULT_PROJECT_DIR, or set CHI2027_FLORENTINE_DIR."
      )
    )
  }

  project_dir <- getwd()
}

preferred_input_file <- file.path(
  project_dir,
  "survey_data_all_participants_1.csv"
)

if (file.exists(preferred_input_file)) {
  input_file <- preferred_input_file
} else {
  input_candidates <- list.files(
    project_dir,
    pattern = "^survey_data_all_participants.*\\.csv$",
    full.names = TRUE,
    ignore.case = TRUE
  )

  if (length(input_candidates) != 1L) {
    stop(
      paste0(
        "Expected one survey CSV but found ", length(input_candidates),
        ". Rename the updated file to survey_data_all_participants_1.csv ",
        "or set preferred_input_file explicitly."
      )
    )
  }

  input_file <- input_candidates[[1L]]
}

output_dir <- file.path(project_dir, "qualitative_analysis_outputs")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

EXPECTED_PARTICIPANTS <- 1:27
EXPECTED_QUALITATIVE_SIGNATURE <- 720210829

cat("Input file:", input_file, "\n")
cat("Output directory:", output_dir, "\n")
```

The semantic signature is calculated only from participant IDs and normalized
open-text responses. This is intentional: fixed qualitative coding must not be
silently applied to another version of the data.

```{r}
#| label: helper-functions

normalize_whitespace <- function(x) {
  x[is.na(x)] <- ""
  trimws(gsub("[[:space:]]+", " ", x))
}

semantic_signature <- function(x) {
  code_points <- utf8ToInt(enc2utf8(paste(x, collapse = "\n")))
  hash <- 0

  for (value in code_points) {
    hash <- (hash * 31 + value) %% 2147483647
  }

  as.numeric(hash)
}

make_assignment <- function(participants, source_question, code) {
  data.frame(
    participant = as.integer(participants),
    source_question = source_question,
    code = code,
    stringsAsFactors = FALSE
  )
}

format_ids <- function(x) {
  paste(sprintf("P%02d", sort(unique(as.integer(x)))), collapse = ", ")
}

collapse_unique <- function(x) {
  paste(sort(unique(x)), collapse = "; ")
}

write_utf8_csv <- function(x, path) {
  connection <- file(path, open = "wt", encoding = "UTF-8")
  on.exit(close(connection), add = TRUE)
  utils::write.csv(x, connection, row.names = FALSE, na = "")
}
```

## Import and data-integrity checks

The raw CSV is read but never modified.

```{r}
#| label: import-survey

survey <- tryCatch(
  utils::read.csv(
    input_file,
    stringsAsFactors = FALSE,
    check.names = FALSE,
    fileEncoding = "UTF-8-BOM",
    na.strings = character(0)
  ),
  error = function(e) {
    utils::read.csv(
      input_file,
      stringsAsFactors = FALSE,
      check.names = FALSE,
      fileEncoding = "UTF-8",
      na.strings = character(0)
    )
  }
)

names(survey) <- sub("^\ufeff", "", names(survey))

participant_col <- "participant"
experience_col <- "What was your experience with the tool?"
trust_col <- "What factors made you trust or distrust the AI's prediction?"

required_columns <- c(participant_col, experience_col, trust_col)
missing_columns <- setdiff(required_columns, names(survey))

if (length(missing_columns) > 0L) {
  stop(
    paste0(
      "Required qualitative columns are missing:\n",
      paste(missing_columns, collapse = "\n")
    )
  )
}

survey$participant <- suppressWarnings(
  as.integer(trimws(as.character(survey[[participant_col]])))
)

if (anyNA(survey$participant)) {
  stop("The participant column contains missing or non-numeric IDs.")
}

if (anyDuplicated(survey$participant)) {
  stop("Participant IDs are not unique.")
}

survey <- survey[order(survey$participant), , drop = FALSE]
row.names(survey) <- NULL

if (!identical(survey$participant, EXPECTED_PARTICIPANTS)) {
  stop(
    paste0(
      "This notebook was coded for participant IDs 1-27. Found: ",
      paste(survey$participant, collapse = ", "),
      ". The responses must be recoded if the sample changes."
    )
  )
}

signature_source <- paste0(
  survey$participant,
  "||",
  normalize_whitespace(survey[[experience_col]]),
  "||",
  normalize_whitespace(survey[[trust_col]])
)

actual_signature <- semantic_signature(signature_source)

if (!identical(actual_signature, EXPECTED_QUALITATIVE_SIGNATURE)) {
  stop(
    paste0(
      "The qualitative responses do not match the coded CSV version.\n",
      "Expected signature: ", EXPECTED_QUALITATIVE_SIGNATURE, "\n",
      "Observed signature: ", actual_signature, "\n",
      "Review and update all coding assignments before rerunning."
    )
  )
}

experience_available <- normalize_whitespace(survey[[experience_col]]) != ""
trust_available <- normalize_whitespace(survey[[trust_col]]) != ""
any_qualitative_text <- experience_available | trust_available

QUALITATIVE_PARTICIPANTS <- survey$participant[any_qualitative_text]
QUALITATIVE_N <- length(QUALITATIVE_PARTICIPANTS)
AVAILABLE_RESPONSE_N <- sum(experience_available) + sum(trust_available)
MISSING_RESPONSE_N <- 2L * nrow(survey) - AVAILABLE_RESPONSE_N
NO_TEXT_PARTICIPANTS <- survey$participant[!any_qualitative_text]

data.frame(
  check = c(
    "Survey participants",
    "Potential open-ended responses",
    "Available open-ended responses",
    "Participants contributing qualitative text",
    "Participants with no qualitative text",
    "Missing qualitative responses",
    "Duplicate participant IDs",
    "Semantic signature"
  ),
  result = c(
    nrow(survey),
    2L * nrow(survey),
    AVAILABLE_RESPONSE_N,
    QUALITATIVE_N,
    format_ids(NO_TEXT_PARTICIPANTS),
    MISSING_RESPONSE_N,
    0L,
    actual_signature
  )
) |>
  knitr::kable(caption = "Data-integrity checks")
```

## Qualitative corpus

The two response fields per participant are converted to long format while
preserving the original wording. Empty fields remain documented in
`responses_long_all`; only non-empty responses enter the analytic corpus. No
spelling or grammar is corrected in the source text. English translations are
provided alongside the two German responses from P27.

```{r}
#| label: create-corpus

responses_long_all <- rbind(
  data.frame(
    participant = survey$participant,
    participant_label = sprintf("P%02d", survey$participant),
    source_question = "tool_experience",
    question_text = experience_col,
    response_text = as.character(survey[[experience_col]]),
    stringsAsFactors = FALSE
  ),
  data.frame(
    participant = survey$participant,
    participant_label = sprintf("P%02d", survey$participant),
    source_question = "trust_factors",
    question_text = trust_col,
    response_text = as.character(survey[[trust_col]]),
    stringsAsFactors = FALSE
  )
)

responses_long_all$normalized_response <- normalize_whitespace(
  responses_long_all$response_text
)
responses_long_all$response_available <-
  responses_long_all$normalized_response != ""
responses_long_all$response_characters <- nchar(
  responses_long_all$normalized_response,
  type = "chars"
)

responses_long_all$translation_en <- ""
responses_long_all$translation_en[
  responses_long_all$participant == 27 &
    responses_long_all$source_question == "tool_experience"
] <- paste0(
  "It helps to recognize what is essential, but it also tempts one not to ",
  "conduct a step-by-step analysis."
)
responses_long_all$translation_en[
  responses_long_all$participant == 27 &
    responses_long_all$source_question == "trust_factors"
] <- "Correlation with age."

responses_long_all$analysis_text <- ifelse(
  responses_long_all$translation_en == "",
  responses_long_all$response_text,
  responses_long_all$translation_en
)

responses_long_all <- responses_long_all[
  order(responses_long_all$participant, responses_long_all$source_question),
  ,
  drop = FALSE
]
row.names(responses_long_all) <- NULL

responses_long <- responses_long_all[
  responses_long_all$response_available,
  ,
  drop = FALSE
]
row.names(responses_long) <- NULL

corpus_summary <- aggregate(
  response_characters ~ source_question,
  data = responses_long,
  FUN = function(x) {
    c(
      n = length(x),
      median = stats::median(x),
      minimum = min(x),
      maximum = max(x)
    )
  }
)

corpus_summary <- data.frame(
  source_question = corpus_summary$source_question,
  corpus_summary$response_characters,
  row.names = NULL
)

knitr::kable(
  corpus_summary,
  caption = "Length of the qualitative responses in characters"
)
```

```{r}
#| label: fig-response-length
#| fig-cap: "Distribution of response lengths for the two open-ended questions. Points represent individual responses; boxes summarize the median and interquartile range."
#| fig-width: 9
#| fig-height: 4.8

responses_long$question_display <- factor(
  responses_long$source_question,
  levels = c("tool_experience", "trust_factors"),
  labels = c("Tool experience", "Trust and distrust factors")
)

question_colors <- c("#0072B2", "#D55E00")
set.seed(2027)

boxplot(
  response_characters ~ question_display,
  data = responses_long,
  horizontal = TRUE,
  outline = FALSE,
  col = grDevices::adjustcolor(question_colors, alpha.f = 0.22),
  border = question_colors,
  xlab = "Response length in characters",
  las = 1
)

stripchart(
  response_characters ~ question_display,
  data = responses_long,
  horizontal = TRUE,
  method = "jitter",
  jitter = 0.12,
  add = TRUE,
  pch = 16,
  col = question_colors
)
```

The trust/distrust prompt elicited longer and more analytically differentiated
responses than the general experience prompt. This asymmetry matters when
interpreting code coverage: a code observed mainly in the trust responses may
partly reflect the structure of the prompt rather than greater substantive
importance.

The complete corpus is displayed below to make the relationship between raw
responses and interpretation inspectable.

```{r}
#| label: display-corpus

knitr::kable(
  responses_long[
    ,
    c(
      "participant_label",
      "source_question",
      "response_text",
      "translation_en"
    )
  ],
  col.names = c(
    "Participant",
    "Question",
    "Original response",
    "English translation when applicable"
  ),
  caption = "Complete qualitative corpus"
)
```

## Codebook

The analysis is semantic: codes describe explicit meanings in responses rather
than latent psychological constructs. A response may receive more than one code
when it contains multiple meaning units.

```{r}
#| label: define-codebook

codebook <- data.frame(
  code = c(
    "ease_or_learnability",
    "interface_friction",
    "perceived_training_value",
    "diagnostic_search_reduction",
    "attention_direction",
    "confidence_or_reassurance",
    "concordance_with_own_interpretation",
    "observed_accuracy",
    "manual_verification",
    "model_fallibility_or_incompleteness",
    "stability_or_agreement_cue",
    "perceived_measurement_capability",
    "training_data_provenance",
    "prior_clinical_ai_experience",
    "reference_case_relevance_or_corroboration",
    "bounded_or_conditional_reliance",
    "low_expertise_drives_reliance",
    "judgment_influence_or_shortcut_risk",
    "unmodeled_clinical_or_acquisition_context"
  ),
  label = c(
    "Ease or learnability",
    "Interface friction",
    "Perceived training value",
    "Diagnostic-search reduction",
    "Attention direction",
    "Confidence or reassurance",
    "Concordance with own interpretation",
    "Observed accuracy",
    "Manual verification",
    "Model fallibility or incompleteness",
    "Stability or agreement cue",
    "Perceived measurement capability",
    "Training-data provenance",
    "Prior clinical-AI experience",
    "Reference-case relevance or corroboration",
    "Bounded or conditional reliance",
    "Low expertise drives reliance",
    "Judgment influence or shortcut risk",
    "Unmodeled clinical or acquisition context"
  ),
  definition = c(
    "The interface is described as easy, readable, or becoming easier after familiarization.",
    "A concrete interaction burden is identified, such as lengthy explanations, slowness, changing axes, or initial confusion.",
    "The tool is described as supporting learning, early-career clinicians, or ECG understanding.",
    "The prediction narrows differentials, shortlists pathologies, or supplies an initial diagnostic direction.",
    "The prediction directs attention to ECG regions, measurements, or decision thresholds.",
    "The support is described as increasing confidence in forming a diagnosis.",
    "Trust follows agreement with the participant's prior diagnosis or visible ECG evidence.",
    "Trust is updated from perceived correctness across cases encountered during the study.",
    "The participant checks AI output against personal interpretation or manual measurements.",
    "Distrust reflects known errors, imperfect output, or incomplete interpretation.",
    "Displayed prediction stability or agreement among models is used as a trust cue.",
    "Trust is based on a belief that AI can measure objective ECG parameters precisely.",
    "Trust depends on the amount, labeling quality, or population diversity of training data.",
    "Prior exposure to hospital ECG AI informs current trust or skepticism.",
    "The perceived relevance, similarity, mismatch, or distracting character of a reference case affects trust or decision formation.",
    "AI is explicitly treated as supportive or advisory rather than definitive evidence.",
    "Limited self-assessed ECG expertise is explicitly given as a reason for trusting AI.",
    "The prediction is said to influence a diagnostic threshold or risk displacing independent reasoning.",
    "Distrust arises because anatomy, acquisition errors, or contextual variables may be outside the model."
  ),
  include_when = c(
    "Explicit ease, readability, or improvement after familiarization.",
    "A specific usability problem is named.",
    "Learning, training, early-career use, or improved understanding is explicit.",
    "Differential reduction, shortlisting, or diagnostic direction is explicit.",
    "An attentional or measurement target is named.",
    "Support is explicitly linked to greater diagnostic confidence.",
    "Trust is conditional on agreement with one's own diagnosis or visible features.",
    "Repeated or high observed correctness is cited.",
    "Counter-checking, measuring, or double-checking is explicit.",
    "Errors, non-perfect performance, or missed findings are mentioned.",
    "A numerical stability/confidence display or model agreement is cited.",
    "Precise analysis of measurable ECG parameters is explicitly cited as a reason for trust.",
    "Training-set size, labels, or diversity is cited.",
    "A hospital or previously used clinical AI system is cited.",
    "A reference patient or ECG explicitly supports, undermines, or distracts from the decision because of its perceived relevance or similarity.",
    "The response explicitly limits reliance, for example by saying that AI is supportive, not definitive, or not fully relied upon.",
    "Poor or limited ECG knowledge is explicitly linked to greater reliance.",
    "The participant explicitly reports AI influence on their own judgment.",
    "Concrete patient or acquisition factors outside the model are listed."
  ),
  exclude_when = c(
    "Generic approval without an ease or learning claim.",
    "Generic dislike without a concrete interaction issue.",
    "General usefulness without a learning or training claim.",
    "Generic helpfulness without a search-space or direction mechanism.",
    "General guidance without an attentional or measurement target.",
    "Trust alone without greater self-confidence.",
    "Agreement between systems rather than with the participant's own judgment.",
    "General model-quality statements without observed study performance.",
    "General skepticism without an active checking action.",
    "Abstract training-data concerns without a failure or incompleteness claim.",
    "Agreement with the participant rather than a displayed system cue.",
    "Observed correctness without a claim about measurement capability.",
    "Observed study accuracy without a training-data claim.",
    "General awareness of AI fallibility without prior clinical-AI experience.",
    "Generic AI agreement without a reference case or a reference case mentioned without an effect on trust or decision-making.",
    "Manual checking alone without an explicit statement that reliance is limited.",
    "Low trust caused by low expertise; this code requires increased reliance.",
    "Helpful guidance without reported influence on independent judgment.",
    "Generic uncertainty without concrete contextual or acquisition factors."
  ),
  stringsAsFactors = FALSE
)

if (anyDuplicated(codebook$code)) {
  stop("Duplicate code names found in the codebook.")
}

knitr::kable(
  codebook,
  caption = "Qualitative codebook and coding boundaries"
)
```

## Explicit coding assignments

Each row created below identifies one participant response and one analytic
code. The assignments are explicit so that every code can be traced to its
source response.

```{r}
#| label: coding-assignments

coding <- do.call(
  rbind,
  list(
    make_assignment(c(1, 3, 15, 19, 24), "tool_experience", "ease_or_learnability"),
    make_assignment(c(3, 9, 19, 25), "tool_experience", "interface_friction"),
    make_assignment(c(4, 16, 17, 23), "tool_experience", "perceived_training_value"),

    make_assignment(c(5, 6), "tool_experience", "diagnostic_search_reduction"),
    make_assignment(21, "trust_factors", "diagnostic_search_reduction"),
    make_assignment(c(11, 27), "tool_experience", "attention_direction"),
    make_assignment(5, "trust_factors", "attention_direction"),
    make_assignment(6, "tool_experience", "confidence_or_reassurance"),

    make_assignment(24, "tool_experience", "concordance_with_own_interpretation"),
    make_assignment(c(1, 4, 6, 14, 24), "trust_factors", "concordance_with_own_interpretation"),
    make_assignment(c(4, 21), "tool_experience", "observed_accuracy"),
    make_assignment(c(2, 4, 6, 9, 21), "trust_factors", "observed_accuracy"),
    make_assignment(c(4, 5, 7, 10, 24), "trust_factors", "manual_verification"),
    make_assignment(c(2, 7, 9, 20, 23), "trust_factors", "model_fallibility_or_incompleteness"),
    make_assignment(c(6, 10, 13, 24), "trust_factors", "stability_or_agreement_cue"),
    make_assignment(11, "trust_factors", "perceived_measurement_capability"),
    make_assignment(c(8, 15), "trust_factors", "training_data_provenance"),
    make_assignment(c(7, 9), "trust_factors", "prior_clinical_ai_experience"),

    make_assignment(c(13, 24, 25), "tool_experience", "reference_case_relevance_or_corroboration"),
    make_assignment(c(3, 17, 18, 24, 25), "trust_factors", "reference_case_relevance_or_corroboration"),

    make_assignment(c(5, 24), "trust_factors", "bounded_or_conditional_reliance"),
    make_assignment(c(12, 16), "trust_factors", "low_expertise_drives_reliance"),
    make_assignment(c(10, 27), "tool_experience", "judgment_influence_or_shortcut_risk"),
    make_assignment(c(19, 24), "trust_factors", "unmodeled_clinical_or_acquisition_context")
  )
)

coding <- unique(coding)
coding <- coding[
  order(coding$participant, coding$source_question, coding$code),
  ,
  drop = FALSE
]
row.names(coding) <- NULL

unknown_codes <- setdiff(coding$code, codebook$code)
if (length(unknown_codes) > 0L) {
  stop(paste("Unknown codes:", paste(unknown_codes, collapse = ", ")))
}

response_keys <- paste(
  responses_long$participant,
  responses_long$source_question,
  sep = "||"
)
coding_keys <- paste(coding$participant, coding$source_question, sep = "||")

if (any(!coding_keys %in% response_keys)) {
  stop("At least one assignment does not point to an existing response.")
}

response_match <- match(coding_keys, response_keys)
coding$response_text <- responses_long$response_text[response_match]
coding$translation_en <- responses_long$translation_en[response_match]
coding$analysis_text <- responses_long$analysis_text[response_match]
coding$participant_label <- sprintf("P%02d", coding$participant)
coding$code_label <- codebook$label[match(coding$code, codebook$code)]
coding$code_definition <- codebook$definition[match(coding$code, codebook$code)]

coding <- coding[
  ,
  c(
    "participant",
    "participant_label",
    "source_question",
    "response_text",
    "translation_en",
    "analysis_text",
    "code",
    "code_label",
    "code_definition"
  )
]

cat("Validated coding assignments:", nrow(coding), "\n")
```

The table makes every interpretive assignment traceable to the original
response.

```{r}
#| label: display-coding

knitr::kable(
  coding[
    ,
    c(
      "participant_label",
      "source_question",
      "code_label",
      "response_text",
      "translation_en"
    )
  ],
  col.names = c(
    "Participant",
    "Question",
    "Code",
    "Original response",
    "English translation when applicable"
  ),
  caption = "Participant-level qualitative coding matrix"
)
```

Responses that do not support a sufficiently specific semantic code are
retained rather than forced into a theme. Generic evaluations indicate valence
but do not identify a mechanism; similarly, a factor named without explaining
how it shaped trust remains analytically ambiguous.

```{r}
#| label: document-uncoded-responses

coded_response_keys <- unique(
  paste(coding$participant, coding$source_question, sep = "||")
)
uncoded_responses <- responses_long[
  !response_keys %in% coded_response_keys,
  ,
  drop = FALSE
]

uncoded_rationales <- data.frame(
  participant = c(2, 7, 8, 12, 14, 18, 20, 26, 26, 27),
  source_question = c(
    rep("tool_experience", 8),
    "trust_factors",
    "trust_factors"
  ),
  exclusion_rationale = c(
    rep(
      "General evaluation without enough detail to identify an analytic mechanism.",
      8
    ),
    "A single abstract noun does not identify the direction or basis of trust.",
    "Age correlation is mentioned, but its relationship to trust or distrust is not explained."
  ),
  stringsAsFactors = FALSE
)

uncoded_keys <- paste(
  uncoded_responses$participant,
  uncoded_responses$source_question,
  sep = "||"
)
rationale_keys <- paste(
  uncoded_rationales$participant,
  uncoded_rationales$source_question,
  sep = "||"
)

if (!setequal(uncoded_keys, rationale_keys)) {
  stop("The documented uncoded responses do not match the uncoded corpus.")
}

uncoded_responses$exclusion_rationale <- uncoded_rationales$exclusion_rationale[
  match(uncoded_keys, rationale_keys)
]

knitr::kable(
  uncoded_responses[
    ,
    c(
      "participant_label",
      "source_question",
      "response_text",
      "translation_en",
      "exclusion_rationale"
    )
  ],
  col.names = c(
    "Participant",
    "Question",
    "Original response",
    "English translation when applicable",
    "Reason no semantic code was assigned"
  ),
  caption = "Non-empty responses retained without a semantic code"
)
```

## Candidate themes

A code may contribute to more than one theme. For example, manual verification
is both a trust-calibration practice and the counterweight in the tension
between guidance and cognitive substitution.

```{r}
#| label: define-themes

theme_definitions <- data.frame(
  theme_id = paste0("T", 1:5),
  theme = c(
    "Prediction as diagnostic navigation",
    "Trust as corroboration and ongoing calibration",
    "Guidance versus cognitive substitution",
    "Reference cases as conditional and contested evidence",
    "Usability, learning curve, and perceived training potential"
  ),
  central_idea = c(
    "Direct predictions reorganized the task by narrowing diagnostic search and directing attention.",
    "Trust was constructed through corroborating cues and updated in light of observed performance and known limitations.",
    "Participants described both active verification and circumstances in which AI could exert disproportionate influence.",
    "Reference cases supported trust when they appeared relevant or similar, but could also distract or undermine trust when they appeared mismatched.",
    "The interface was generally learnable but contained specific friction, alongside perceived learning value."
  ),
  boundary = c(
    "This describes perceived mechanisms, not logged gaze or search behavior.",
    "The prompted trust question maps trust cues but does not estimate population prevalence.",
    "The study does not demonstrate automation bias because displayed predictions were correct and no initial unaided judgment was recorded.",
    "The open responses contain both corroborating and negative reference-case accounts, but cannot establish an average behavioral effect.",
    "Educational effectiveness, retention, and transfer were not measured; only perceived potential is supported."
  ),
  design_implication = c(
    "Keep support concise while preserving access to evidence behind the narrowed search space.",
    "Expose provenance and uncertainty carefully and support manual corroboration rather than treating trust as a single score.",
    "Consider eliciting an initial assessment or showing support on request, particularly for less experienced users.",
    "Align corresponding ECG leads, annotate discriminating features, and explain why a case was retrieved.",
    "Use progressive disclosure, stable axes, concise explanations, and separate training from point-of-care modes."
  ),
  stringsAsFactors = FALSE
)

theme_map <- rbind(
  data.frame(
    theme_id = "T1",
    code = c(
      "diagnostic_search_reduction",
      "attention_direction",
      "confidence_or_reassurance"
    ),
    stringsAsFactors = FALSE
  ),
  data.frame(
    theme_id = "T2",
    code = c(
      "concordance_with_own_interpretation",
      "observed_accuracy",
      "manual_verification",
      "model_fallibility_or_incompleteness",
      "stability_or_agreement_cue",
      "perceived_measurement_capability",
      "training_data_provenance",
      "prior_clinical_ai_experience",
      "bounded_or_conditional_reliance",
      "unmodeled_clinical_or_acquisition_context"
    ),
    stringsAsFactors = FALSE
  ),
  data.frame(
    theme_id = "T3",
    code = c(
      "manual_verification",
      "model_fallibility_or_incompleteness",
      "bounded_or_conditional_reliance",
      "low_expertise_drives_reliance",
      "judgment_influence_or_shortcut_risk"
    ),
    stringsAsFactors = FALSE
  ),
  data.frame(
    theme_id = "T4",
    code = "reference_case_relevance_or_corroboration",
    stringsAsFactors = FALSE
  ),
  data.frame(
    theme_id = "T5",
    code = c(
      "ease_or_learnability",
      "interface_friction",
      "perceived_training_value"
    ),
    stringsAsFactors = FALSE
  )
)

if (any(!theme_map$code %in% codebook$code)) {
  stop("Theme map contains a code absent from the codebook.")
}

coding_theme <- merge(
  coding,
  theme_map,
  by = "code",
  all.x = FALSE,
  all.y = FALSE,
  sort = FALSE
)
coding_theme <- unique(coding_theme)
coding_theme$theme <- theme_definitions$theme[
  match(coding_theme$theme_id, theme_definitions$theme_id)
]

coding_theme <- coding_theme[
  ,
  c(
    "participant",
    "participant_label",
    "source_question",
    "response_text",
    "translation_en",
    "analysis_text",
    "code",
    "code_label",
    "code_definition",
    "theme_id",
    "theme"
  )
]

knitr::kable(
  merge(theme_map, theme_definitions[, c("theme_id", "theme")], by = "theme_id"),
  col.names = c("Theme ID", "Code", "Theme"),
  caption = "Mapping from codes to candidate themes"
)
```

## Descriptive coverage

Counts refer to unique participants represented in a code or theme. They are
used as descriptive corpus information, not as inferential tests, prevalence
estimates, or measures of thematic importance. Counts overlap because one
participant can contribute to several codes and themes.

```{r}
#| label: code-theme-summaries

unique_code_participants <- unique(coding[, c("code", "participant")])

code_n <- aggregate(
  participant ~ code,
  data = unique_code_participants,
  FUN = length
)
names(code_n)[names(code_n) == "participant"] <- "n_participants"

code_ids <- aggregate(
  participant ~ code,
  data = unique_code_participants,
  FUN = format_ids
)
names(code_ids)[names(code_ids) == "participant"] <- "participant_ids"

code_summary <- merge(codebook, code_n, by = "code", all.x = TRUE, sort = FALSE)
code_summary <- merge(code_summary, code_ids, by = "code", all.x = TRUE, sort = FALSE)
code_summary$n_participants[is.na(code_summary$n_participants)] <- 0L
code_summary$participant_ids[is.na(code_summary$participant_ids)] <- ""
code_summary$coverage_percent <- round(
  100 * code_summary$n_participants / QUALITATIVE_N,
  1
)
code_summary <- code_summary[
  match(codebook$code, code_summary$code),
  ,
  drop = FALSE
]

unique_theme_participants <- unique(
  coding_theme[, c("theme_id", "theme", "participant")]
)

theme_n <- aggregate(
  participant ~ theme_id + theme,
  data = unique_theme_participants,
  FUN = length
)
names(theme_n)[names(theme_n) == "participant"] <- "n_participants"

theme_ids <- aggregate(
  participant ~ theme_id + theme,
  data = unique_theme_participants,
  FUN = format_ids
)
names(theme_ids)[names(theme_ids) == "participant"] <- "participant_ids"

theme_summary <- merge(
  theme_definitions,
  theme_n,
  by = c("theme_id", "theme"),
  all.x = TRUE,
  sort = FALSE
)
theme_summary <- merge(
  theme_summary,
  theme_ids,
  by = c("theme_id", "theme"),
  all.x = TRUE,
  sort = FALSE
)
theme_summary$coverage_percent <- round(
  100 * theme_summary$n_participants / QUALITATIVE_N,
  1
)
theme_summary <- theme_summary[
  match(theme_definitions$theme_id, theme_summary$theme_id),
  ,
  drop = FALSE
]

knitr::kable(
  code_summary[
    ,
    c("label", "n_participants", "coverage_percent", "participant_ids")
  ],
  col.names = c("Code", "Participants", "Coverage (%)", "Participant IDs"),
  caption = "Code coverage using unique participants"
)
```

```{r}
#| label: fig-code-coverage
#| fig-cap: "Participant coverage of the 19 semantic codes. Counts are overlapping and descriptive."
#| fig-width: 9
#| fig-height: 7.5

code_plot_order <- order(
  code_summary$n_participants,
  code_summary$label
)

old_par <- par(no.readonly = TRUE)
par(mar = c(4.5, 15.5, 1.5, 2))

dotchart(
  code_summary$n_participants[code_plot_order],
  labels = code_summary$label[code_plot_order],
  xlim = c(0, max(code_summary$n_participants) + 1.2),
  pch = 19,
  color = "#0072B2",
  lcolor = "#B8C2CC",
  xlab = "Unique participants represented in code",
  cex = 0.82
)

text(
  x = code_summary$n_participants[code_plot_order] + 0.22,
  y = seq_along(code_plot_order),
  labels = code_summary$n_participants[code_plot_order],
  pos = 4,
  cex = 0.82
)

par(old_par)
```

### Prompt-conditioned coding patterns

Because each question invited a different kind of reflection, code coverage is
also decomposed by prompt. This helps distinguish an analytic pattern from a
question-format effect.

```{r}
#| label: code-by-question

unique_code_question_participants <- unique(
  coding[, c("code", "source_question", "participant")]
)

code_question_counts <- aggregate(
  participant ~ code + source_question,
  data = unique_code_question_participants,
  FUN = length
)
names(code_question_counts)[names(code_question_counts) == "participant"] <-
  "n_participants"

code_question_matrix <- matrix(
  0L,
  nrow = nrow(codebook),
  ncol = 2L,
  dimnames = list(
    codebook$code,
    c("tool_experience", "trust_factors")
  )
)

matrix_rows <- match(code_question_counts$code, rownames(code_question_matrix))
matrix_cols <- match(
  code_question_counts$source_question,
  colnames(code_question_matrix)
)
code_question_matrix[cbind(matrix_rows, matrix_cols)] <-
  code_question_counts$n_participants

code_question_summary <- data.frame(
  code = codebook$code,
  code_label = codebook$label,
  tool_experience = code_question_matrix[, "tool_experience"],
  trust_factors = code_question_matrix[, "trust_factors"],
  total_unique_participants = code_summary$n_participants,
  stringsAsFactors = FALSE,
  row.names = NULL
)

knitr::kable(
  code_question_summary[
    ,
    c("code_label", "tool_experience", "trust_factors", "total_unique_participants")
  ],
  col.names = c(
    "Code",
    "Tool-experience prompt",
    "Trust prompt",
    "Unique participants overall"
  ),
  caption = "Code coverage separated by open-ended question"
)
```

```{r}
#| label: fig-code-question-heatmap
#| fig-cap: "Prompt-conditioned code coverage. Each cell reports the number of unique participants contributing that code within the question."
#| fig-width: 9
#| fig-height: 8.5

display_code_matrix <- code_question_matrix[
  rev(seq_len(nrow(code_question_matrix))),
  ,
  drop = FALSE
]
heatmap_matrix <- t(display_code_matrix)
maximum_cell <- max(heatmap_matrix)
heatmap_colors <- hcl.colors(maximum_cell + 1L, "Blues 3")

old_par <- par(no.readonly = TRUE)
par(mar = c(6, 17.5, 1.5, 2))

image(
  x = seq_len(ncol(display_code_matrix)),
  y = seq_len(nrow(display_code_matrix)),
  z = heatmap_matrix,
  col = heatmap_colors,
  breaks = seq(-0.5, maximum_cell + 0.5, by = 1),
  axes = FALSE,
  xlab = "",
  ylab = ""
)

axis(
  1,
  at = 1:2,
  labels = c("Tool experience", "Trust / distrust factors"),
  las = 2
)
axis(
  2,
  at = seq_len(nrow(display_code_matrix)),
  labels = rev(codebook$label),
  las = 1,
  cex.axis = 0.78
)

abline(v = 1.5, col = "white", lwd = 2)
abline(
  h = seq(1.5, nrow(display_code_matrix) - 0.5, by = 1),
  col = "white",
  lwd = 1
)

cell_locations <- expand.grid(
  x = seq_len(ncol(display_code_matrix)),
  y = seq_len(nrow(display_code_matrix))
)
cell_values <- as.vector(heatmap_matrix)

text(
  cell_locations$x,
  cell_locations$y,
  labels = cell_values,
  col = ifelse(cell_values > maximum_cell / 2, "white", "black"),
  font = 2
)

par(old_par)
```

The split confirms that usability, learning, and task-navigation mechanisms
were expressed mainly in answers about the tool experience, whereas almost all
calibration mechanisms arose in the explicitly prompted trust question. The
reference-case code is the exception: it appears in both prompts, suggesting
that comparison with a retrieved case crossed the boundary between general
experience and trust formation.

```{r}
#| label: theme-summary-table

knitr::kable(
  theme_summary[
    ,
    c("theme_id", "theme", "n_participants", "coverage_percent", "participant_ids")
  ],
  col.names = c("ID", "Theme", "Participants", "Coverage (%)", "Participant IDs"),
  caption = "Overlapping descriptive coverage of candidate themes"
)
```

```{r}
#| label: fig-theme-coverage
#| fig-cap: "Participant coverage of the five candidate themes. Counts overlap and are not prevalence estimates."
#| fig-width: 9
#| fig-height: 5.5

plot_order <- rev(seq_len(nrow(theme_summary)))
plot_labels <- vapply(
  theme_summary$theme,
  function(x) paste(strwrap(x, width = 38), collapse = "\n"),
  character(1)
)

old_par <- par(no.readonly = TRUE)
par(mar = c(4.5, 17, 1.5, 2))

bar_midpoints <- barplot(
  theme_summary$n_participants[plot_order],
  names.arg = plot_labels[plot_order],
  horiz = TRUE,
  las = 1,
  xlim = c(0, QUALITATIVE_N + 1),
  col = "#2C7FB8",
  border = NA,
  axes = FALSE,
  xlab = "Unique participants represented in theme"
)

text(
  x = theme_summary$n_participants[plot_order] + 0.35,
  y = bar_midpoints,
  labels = paste0(theme_summary$n_participants[plot_order], "/", QUALITATIVE_N),
  adj = 0,
  cex = 0.9
)

axis(1, at = seq(0, QUALITATIVE_N, by = 5))
par(old_par)
```

## Relational and configurational analysis

Theme counts show how broadly each idea appears, but they do not show how ideas
combine within participants. The following analyses retain participant identity
and examine configurations, overlaps, and contrasting accounts.

### Participant–theme configurations

```{r}
#| label: participant-theme-matrix

participant_theme_matrix <- matrix(
  0L,
  nrow = QUALITATIVE_N,
  ncol = nrow(theme_definitions),
  dimnames = list(
    sprintf("P%02d", QUALITATIVE_PARTICIPANTS),
    theme_definitions$theme_id
  )
)

unique_memberships <- unique(
  coding_theme[, c("participant_label", "theme_id")]
)
membership_rows <- match(
  unique_memberships$participant_label,
  rownames(participant_theme_matrix)
)
membership_cols <- match(
  unique_memberships$theme_id,
  colnames(participant_theme_matrix)
)
participant_theme_matrix[cbind(membership_rows, membership_cols)] <- 1L

participant_theme_configuration <- data.frame(
  participant_label = rownames(participant_theme_matrix),
  n_themes = rowSums(participant_theme_matrix),
  themes = apply(
    participant_theme_matrix,
    1,
    function(x) paste(names(x)[x == 1L], collapse = "; ")
  ),
  stringsAsFactors = FALSE,
  row.names = NULL
)

knitr::kable(
  participant_theme_configuration,
  col.names = c("Participant", "Number of themes", "Theme configuration"),
  caption = "Theme configurations represented by each participant"
)
```

```{r}
#| label: fig-participant-theme-map
#| fig-cap: "Participant–theme configuration map. Colored cells indicate that at least one response from the participant contributed to the theme."
#| fig-width: 8.5
#| fig-height: 8

display_presence <- participant_theme_matrix[
  rev(seq_len(nrow(participant_theme_matrix))),
  ,
  drop = FALSE
]
presence_image <- t(display_presence)

old_par <- par(no.readonly = TRUE)
par(mar = c(5, 5, 1.5, 2))

image(
  x = seq_len(ncol(display_presence)),
  y = seq_len(nrow(display_presence)),
  z = presence_image,
  col = c("#F2F2F2", "#009E73"),
  breaks = c(-0.5, 0.5, 1.5),
  axes = FALSE,
  xlab = "Candidate theme",
  ylab = "Participant"
)

axis(1, at = 1:5, labels = theme_definitions$theme_id)
axis(
  2,
  at = seq_len(nrow(display_presence)),
  labels = rev(rownames(participant_theme_matrix)),
  las = 1,
  cex.axis = 0.78
)
abline(v = seq(1.5, 4.5, by = 1), col = "white", lwd = 1.5)
abline(h = seq(1.5, nrow(display_presence) - 0.5, by = 1), col = "white")

par(old_par)
```

The map shows that themes are not mutually exclusive participant types. Several
participants contributed simultaneously to trust calibration and its possible
failure modes, while others provided narrow but analytically distinct evidence
about navigation, reference cases, or usability. P26 provided two very brief
responses, but neither contained enough semantic detail to support a specific
code; the zero-theme row preserves that case rather than forcing an
interpretation.

### Theme co-occurrence and similarity

The co-occurrence matrix counts participants represented in two themes. Jaccard
similarity divides this intersection by the number represented in either theme,
thereby accounting for the different sizes of themes:

\[
J(A,B)=\frac{|A\cap B|}{|A\cup B|}.
\]

```{r}
#| label: theme-overlap-analysis

theme_intersections <- crossprod(participant_theme_matrix)
theme_sizes <- diag(theme_intersections)

theme_jaccard <- matrix(
  0,
  nrow = nrow(theme_intersections),
  ncol = ncol(theme_intersections),
  dimnames = dimnames(theme_intersections)
)

for (row_index in seq_len(nrow(theme_jaccard))) {
  for (column_index in seq_len(ncol(theme_jaccard))) {
    intersection <- theme_intersections[row_index, column_index]
    union <- theme_sizes[row_index] + theme_sizes[column_index] - intersection
    theme_jaccard[row_index, column_index] <- intersection / union
  }
}

theme_pairs <- utils::combn(theme_definitions$theme_id, 2)
theme_overlap_pairs <- data.frame(
  theme_a = theme_pairs[1, ],
  theme_b = theme_pairs[2, ],
  shared_participants = mapply(
    function(a, b) theme_intersections[a, b],
    theme_pairs[1, ],
    theme_pairs[2, ]
  ),
  participant_ids = mapply(
    function(a, b) {
      format_ids(
        QUALITATIVE_PARTICIPANTS[
          participant_theme_matrix[, a] == 1L &
            participant_theme_matrix[, b] == 1L
        ]
      )
    },
    theme_pairs[1, ],
    theme_pairs[2, ]
  ),
  jaccard_similarity = mapply(
    function(a, b) theme_jaccard[a, b],
    theme_pairs[1, ],
    theme_pairs[2, ]
  ),
  stringsAsFactors = FALSE
)
theme_overlap_pairs$jaccard_similarity <- round(
  theme_overlap_pairs$jaccard_similarity,
  2
)
theme_overlap_pairs <- theme_overlap_pairs[
  order(
    -theme_overlap_pairs$jaccard_similarity,
    -theme_overlap_pairs$shared_participants
  ),
  ,
  drop = FALSE
]
row.names(theme_overlap_pairs) <- NULL

knitr::kable(
  theme_overlap_pairs,
  col.names = c(
    "Theme A",
    "Theme B",
    "Shared participants",
    "Participant IDs",
    "Jaccard similarity"
  ),
  digits = 2,
  caption = "Pairwise overlap between candidate themes"
)
```

```{r}
#| label: fig-theme-jaccard
#| fig-cap: "Jaccard similarity between candidate themes based on participant membership. The diagonal equals 1 by definition."
#| fig-width: 7
#| fig-height: 6.5

display_jaccard <- theme_jaccard[
  ,
  rev(seq_len(ncol(theme_jaccard))),
  drop = FALSE
]
jaccard_colors <- hcl.colors(101, "YlGnBu")

old_par <- par(no.readonly = TRUE)
par(mar = c(5, 5, 1.5, 2))

image(
  x = 1:5,
  y = 1:5,
  z = display_jaccard,
  col = jaccard_colors,
  breaks = seq(0, 1, length.out = 102),
  zlim = c(0, 1),
  axes = FALSE,
  xlab = "Theme",
  ylab = "Theme"
)

axis(1, at = 1:5, labels = theme_definitions$theme_id)
axis(2, at = 1:5, labels = rev(theme_definitions$theme_id), las = 1)
abline(v = seq(1.5, 4.5, by = 1), col = "white", lwd = 1.5)
abline(h = seq(1.5, 4.5, by = 1), col = "white", lwd = 1.5)

jaccard_locations <- expand.grid(x = 1:5, y = 1:5)
jaccard_values <- as.vector(display_jaccard)
text(
  jaccard_locations$x,
  jaccard_locations$y,
  labels = sprintf("%.2f", jaccard_values),
  col = ifelse(jaccard_values >= 0.45, "white", "black"),
  font = 2,
  cex = 0.9
)

par(old_par)
```

The strongest non-diagonal overlap is between **T2 (trust calibration)** and
**T3 (guidance versus cognitive substitution)**: nine participants contribute
to both, with a Jaccard similarity of 0.43. This is analytically important. The
risk of disproportionate influence does not appear as a separate rejection of
AI; it is embedded within accounts of monitoring, checking, and recalibrating
trust. T1 and T2 share three participants, indicating that task navigation and
trust calibration sometimes co-occur but are not interchangeable mechanisms.

### Epistemic sources shaping trust and reliance

The trust-related codes can be reorganized by the source of evidence on which a
participant based trust, distrust, or reliance. This second analytic lens cuts
across the five themes and distinguishes clinical corroboration from observed
system performance, model metadata, contextual knowledge, reference-case
comparison, and expertise-linked vulnerability.

```{r}
#| label: trust-evidence-sources

trust_evidence_map <- data.frame(
  evidence_source = c(
    "Participant's own clinical assessment",
    "Participant's own clinical assessment",
    "Observed system behavior",
    "Observed system behavior",
    "Model or interface information",
    "Model or interface information",
    "Model or interface information",
    "Prior experience or contextual knowledge",
    "Prior experience or contextual knowledge",
    "Retrieved-case comparison",
    "Explicit limits on reliance",
    "Expertise-linked vulnerability",
    "Expertise-linked vulnerability"
  ),
  code = c(
    "concordance_with_own_interpretation",
    "manual_verification",
    "observed_accuracy",
    "model_fallibility_or_incompleteness",
    "stability_or_agreement_cue",
    "perceived_measurement_capability",
    "training_data_provenance",
    "prior_clinical_ai_experience",
    "unmodeled_clinical_or_acquisition_context",
    "reference_case_relevance_or_corroboration",
    "bounded_or_conditional_reliance",
    "low_expertise_drives_reliance",
    "judgment_influence_or_shortcut_risk"
  ),
  stringsAsFactors = FALSE
)

evidence_coding <- merge(
  unique(coding[, c("participant", "participant_label", "code")]),
  trust_evidence_map,
  by = "code",
  all = FALSE,
  sort = FALSE
)
evidence_coding$code_label <- codebook$label[
  match(evidence_coding$code, codebook$code)
]

unique_evidence_participants <- unique(
  evidence_coding[, c("evidence_source", "participant")]
)

evidence_n <- aggregate(
  participant ~ evidence_source,
  data = unique_evidence_participants,
  FUN = length
)
names(evidence_n)[2] <- "n_participants"

evidence_ids <- aggregate(
  participant ~ evidence_source,
  data = unique_evidence_participants,
  FUN = format_ids
)
names(evidence_ids)[2] <- "participant_ids"

evidence_codes <- aggregate(
  code_label ~ evidence_source,
  data = unique(evidence_coding[, c("evidence_source", "code_label")]),
  FUN = collapse_unique
)
names(evidence_codes)[2] <- "contributing_codes"

evidence_source_summary <- merge(
  merge(evidence_n, evidence_ids, by = "evidence_source"),
  evidence_codes,
  by = "evidence_source"
)
evidence_source_summary <- evidence_source_summary[
  order(-evidence_source_summary$n_participants, evidence_source_summary$evidence_source),
  ,
  drop = FALSE
]
row.names(evidence_source_summary) <- NULL

knitr::kable(
  evidence_source_summary,
  col.names = c(
    "Evidence source",
    "Participants",
    "Participant IDs",
    "Contributing codes"
  ),
  caption = "Sources shaping trust, distrust, or reliance"
)
```

```{r}
#| label: fig-trust-evidence-sources
#| fig-cap: "Participant coverage of the non-exclusive evidence sources shaping trust, distrust, or reliance."
#| fig-width: 9
#| fig-height: 5.8

evidence_plot_order <- order(evidence_source_summary$n_participants)
evidence_labels <- vapply(
  evidence_source_summary$evidence_source,
  function(x) paste(strwrap(x, width = 35), collapse = "\n"),
  character(1)
)
evidence_colors <- hcl.colors(
  nrow(evidence_source_summary),
  "Dark 3"
)

old_par <- par(no.readonly = TRUE)
par(mar = c(4.5, 15.5, 1.5, 2))

evidence_midpoints <- barplot(
  evidence_source_summary$n_participants[evidence_plot_order],
  names.arg = evidence_labels[evidence_plot_order],
  horiz = TRUE,
  las = 1,
  xlim = c(0, max(evidence_source_summary$n_participants) + 1.2),
  col = evidence_colors[evidence_plot_order],
  border = NA,
  axes = FALSE,
  xlab = "Unique participants"
)
axis(1, at = 0:max(evidence_source_summary$n_participants))
text(
  evidence_source_summary$n_participants[evidence_plot_order] + 0.22,
  evidence_midpoints,
  labels = evidence_source_summary$n_participants[evidence_plot_order],
  adj = 0,
  cex = 0.9
)

par(old_par)
```

Two sources have the broadest coverage: participants' own clinical assessment
and observed system behavior, each represented by eight participants. This
supports a view of trust as empirical corroboration rather than passive
acceptance. Model/interface information and retrieved-case comparison are also
salient (seven and six participants, respectively), but neither is the sole
basis for trust. The
four expertise-linked cases are
qualitatively important despite their lower coverage because they identify a
specific vulnerability: support may receive more weight when independent ECG
knowledge is limited or when the prediction shapes the reasoning process.
Two participants explicitly bounded their reliance by describing the output as
guidance or supportive information rather than definitive evidence.

### Contrasting and internally ambivalent cases

The following cases resist a simple trusted/distrusted or helpful/unhelpful
classification. They expose mechanisms and boundary conditions that are lost in
aggregate counts.

```{r}
#| label: contrasting-cases

contrast_cases <- data.frame(
  participant = c(
    "P05", "P09", "P10", "P18", "P19", "P24", "P25", "P27"
  ),
  analytic_tension = c(
    "Guidance with retained verification",
    "Prior distrust with local recalibration",
    "Confidence cue with active skepticism and influence",
    "Reference support conditional on relevance",
    "Learnability with contextual model skepticism",
    "Multi-cue trust with explicitly bounded reliance",
    "Helpful prediction with a distracting comparator",
    "Attention support with shortcut risk"
  ),
  evidence = c(
    "The prediction guided the participant, who nevertheless zoomed in and manually calculated a borderline PR interval.",
    "Hospital ECG AI was described as almost always wrong, whereas the study system was described as quite accurate.",
    "Confidence numbers increased trust; the participant also searched for errors, double-checked manually, and reported influence on judgment.",
    "The reference point supported trust when relevant findings were present and distrust when they were absent.",
    "Initial interface confusion diminished with use, while anatomical and acquisition factors remained reasons for distrust.",
    "Visible ECG features, the consistency score, and a similar reference case increased trust, while borderline measurements, noise, and conflict with personal interpretation reduced it; AI remained supportive rather than definitive.",
    "The direct prediction helped, but the reference patient distracted and dissimilar ECGs reduced trust.",
    "The tool helped identify essential features but also tempted the participant to omit step-by-step analysis."
  ),
  analytical_significance = c(
    "Navigation did not necessarily eliminate independent checking.",
    "Trust was updated for the local system rather than transferred wholesale from previous AI experience.",
    "Trust, skepticism, verification, and influence can coexist within one encounter.",
    "Retrieved examples functioned as conditional evidence, not as an intrinsically trusted explanation.",
    "Usability acceptance and epistemic trust are separable judgments.",
    "Trust is assembled from multiple cues and can coexist with an explicit limit on reliance.",
    "Prediction support and case-based support can be evaluated differently within the same experience.",
    "Attention direction and cognitive substitution can be two sides of the same mechanism."
  ),
  stringsAsFactors = FALSE
)

knitr::kable(
  contrast_cases,
  col.names = c(
    "Participant",
    "Analytic tension",
    "Evidence across responses",
    "Analytical significance"
  ),
  caption = "Contrasting cases and boundary conditions"
)
```

Together, these cases sharpen the central interpretation: reliance cannot be
placed on a single continuum from trust to distrust. Participants combined
task guidance, clinical checking, system-level cues, prior expectations, and
local judgments of relevance in different ways.

## Exemplar quotations and source verification

The quoted text remains verbatim, including grammatical and spelling errors.
Every excerpt is checked as an exact substring of the relevant raw response;
rendering stops if any quotation has been altered or attributed to the wrong
participant.

```{r}
#| label: verify-quotes

quotes <- data.frame(
  theme_id = c(
    "T1", "T1", "T1", "T1", "T1",
    "T2", "T2", "T2", "T2", "T2", "T2",
    "T3", "T3", "T3", "T3", "T3",
    "T4", "T4", "T4", "T4", "T4",
    "T5", "T5", "T5", "T5"
  ),
  participant = c(
    5, 6, 11, 21, 27,
    4, 10, 9, 23, 24, 11,
    12, 10, 16, 27, 24,
    13, 17, 24, 25, 25,
    3, 9, 19, 23
  ),
  source_question = c(
    "tool_experience", "tool_experience", "tool_experience", "trust_factors", "tool_experience",
    "trust_factors", "trust_factors", "trust_factors", "trust_factors", "trust_factors", "trust_factors",
    "trust_factors", "tool_experience", "trust_factors", "tool_experience", "trust_factors",
    "tool_experience", "trust_factors", "trust_factors", "tool_experience", "trust_factors",
    "tool_experience", "tool_experience", "tool_experience", "tool_experience"
  ),
  excerpt = c(
    "AI prediction does narrow down the possible differentials",
    "helped to find a direction and made more confident in finding the right diagnosis.",
    "helpful, to get your attention to specific parts of the ECG due to the AI prediction.",
    "shortlist the pathologies.",
    "Es hilft, wesentliches zu erkennen",
    "I counter checked the AI's prediction with my experience.",
    "i double checked everything manually anyway",
    "The AI prediction in the hospital is almost always wrong",
    "it omitted minor details like pacemaker spikes",
    "I therefore treated the AI as supportive information rather than definitive evidence.",
    "specific measureable parameters that an AI can analyse very precise.",
    "my bad knowledge of ECGs made me trust the predictions",
    "I felt it influenced my judgement while looking at the ecg.",
    "I trusted AI predictions as I am not very pro at reading ecg",
    "verleitet allerdings auch dazu, keine schritt für schritt Analyse durchzuführen",
    "I therefore treated the AI as supportive information rather than definitive evidence.",
    "It was nice to have references to help in making a decision.",
    "When AI predictions matched the reference ECG, I trusted it.",
    "especially when the consistency score was high and the reference patient showed a similar pattern.",
    "The AI diagnostic prediction helped, but the reference patient was distracting",
    "I distrusted the AI's prediction when the patient and reference patient's ECG's were unalike.",
    "Very easy to use, very long explanation",
    "it's just a bit slow and the switches between the x-axis size were a bit confusing in the beginning",
    "it was a bit confusing to start, however after getting used to it, i found it easier to use.",
    "instructive and helped revise my branch blocks"
  ),
  stringsAsFactors = FALSE
)

quotes$participant_label <- sprintf("P%02d", quotes$participant)
quotes$theme <- theme_definitions$theme[
  match(quotes$theme_id, theme_definitions$theme_id)
]

quote_keys <- paste(quotes$participant, quotes$source_question, sep = "||")
quote_response_index <- match(quote_keys, response_keys)

if (anyNA(quote_response_index)) {
  stop("At least one quotation does not point to an existing response.")
}

quotes$full_response <- responses_long$response_text[quote_response_index]
quotes$translation_en <- responses_long$translation_en[quote_response_index]
quotes$verified <- mapply(
  function(excerpt, full_response) {
    grepl(excerpt, full_response, fixed = TRUE)
  },
  quotes$excerpt,
  quotes$full_response,
  USE.NAMES = FALSE
)

if (any(!quotes$verified)) {
  stop("At least one exemplar quotation failed source verification.")
}

knitr::kable(
  quotes[
    ,
    c("theme_id", "participant_label", "excerpt", "translation_en", "verified")
  ],
  col.names = c(
    "Theme",
    "Participant",
    "Verified verbatim excerpt",
    "English translation when applicable",
    "Verified"
  ),
  caption = "Exemplar quotations verified against the raw CSV"
)
```

## Thematic results and interpretation

### Theme 1 — Prediction as diagnostic navigation

**Coverage:** `r theme_summary$n_participants[theme_summary$theme_id == "T1"]` of
`r QUALITATIVE_N` qualitative contributors.

Participants described direct prediction as an information-compression
mechanism: it reduced the set of diagnostic candidates, supplied an initial
direction, and indicated what to inspect. The value participants described was
therefore not limited to receiving a proposed answer; the prediction also
reorganized the diagnostic task.

> “AI prediction does narrow down the possible differentials” — P05

> “helped to find a direction and made more confident in finding the right diagnosis.” — P06

> “helpful, to get your attention to specific parts of the ECG due to the AI prediction.” — P11

> “It helps to recognize what is essential [...]” — P27, translated from German

P27's account sharpens the mechanism: identifying what is essential can be
helpful, but the same attentional compression can also discourage a systematic
step-by-step reading. Diagnostic navigation is therefore related to, but not
identical with, safe or appropriate reliance.

**Boundary:** these are self-reported mechanisms. The study did not log visual
attention or diagnostic-search behavior, so the data cannot establish that
attention objectively changed.

**Design implication:** concise predictions can help users manage a complex
task, but the interface should preserve access to the evidence supporting the
narrowed search space.

### Theme 2 — Trust as corroboration and ongoing calibration

**Coverage:** `r theme_summary$n_participants[theme_summary$theme_id == "T2"]` of
`r QUALITATIVE_N` qualitative contributors.

Trust was described as situated and revisable rather than as a fixed attitude.
Participants referred to agreement with their own ECG reading, perceived
accuracy across cases, manual checking, the displayed stability or agreement
signal, training-data provenance, previous experience with clinical AI, and
known contextual limitations.

> “I counter checked the AI's prediction with my experience. It was 100% accurate.” — P04

> “The confidence numbers helped me trust the prediction [...] i double checked everything manually anyway” — P10

> “The AI prediction in the hospital is almost always wrong [...] This one was quite accurate” — P09

> “it omitted minor details like pacemaker spikes” — P23

> “I therefore treated the AI as supportive information rather than definitive evidence.” — P24

> “specific measureable parameters that an AI can analyse very precise.” — P11

The stronger interpretation is therefore not simply that participants “trusted
AI.” They negotiated trust through several forms of corroboration, updated it
using observed system behavior, and in some cases placed an explicit boundary
around reliance. P24 is especially informative because trust increased when
clinical features, the consistency display, and the reference case converged,
but decreased with borderline measurements, noise, or conflict with personal
interpretation. P11 contributed a different competence-based cue: the belief
that an AI is well suited to precise analysis of measurable ECG parameters.

**Boundary:** because the question directly prompted respondents to discuss
trust, coverage counts should not be interpreted as spontaneous prevalence in a
broader population.

**Design implication:** interfaces should support verification and expose
provenance and uncertainty carefully rather than representing trust as a single
scalar response to a confidence-like number.

### Theme 3 — Guidance versus cognitive substitution

**Coverage:** `r theme_summary$n_participants[theme_summary$theme_id == "T3"]` of
`r QUALITATIVE_N` qualitative contributors.

The same efficiency that made AI helpful also created a potential
vulnerability. Five participants described manual verification and five
identified model errors or incomplete interpretations. Two explicitly
connected limited ECG expertise to increased trust, while two described direct
influence or a shortcut risk. Two participants also explicitly bounded their
reliance, showing that verification, influence, and restraint are not mutually
exclusive positions.

> “my bad knowledge of ECGs made me trust the predictions” — P12

> “I felt it influenced my judgement while looking at the ecg. Especially when deciding if measured times were pathological.” — P10

> “I trusted AI predictions as I am not very pro at reading ecg” — P16

> “It helps to recognize what is essential, but it also tempts one not to conduct a step-by-step analysis.” — P27, translated from German

> “I therefore treated the AI as supportive information rather than definitive evidence.” — P24

The defensible claim is a **perceived tension between active verification and
cognitive substitution**. The responses suggest conditions under which support
might displace independent reasoning, but they do not demonstrate behavioral
overreliance or automation bias.

**Design implication:** a future interface could elicit an initial assessment
before revealing support, make predictions available on request, or adapt the
timing of support to user expertise.

### Theme 4 — Reference cases as conditional and contested evidence

**Coverage:** `r theme_summary$n_participants[theme_summary$theme_id == "T4"]` of
`r QUALITATIVE_N` qualitative contributors.

Six participants discussed the relevance of a reference case or reference
point. Their accounts were not uniformly positive: comparison supported trust
when the ECGs appeared meaningfully similar, but mismatch or distraction could
undermine it.

> “It was nice to have references to help in making a decision.” — P13

> “When AI predictions matched the reference ECG, I trusted it.” — P17

> “The AI diagnostic prediction helped, but the reference patient was distracting” — P25

> “I distrusted the AI's prediction when the patient and reference patient's ECG's were unalike.” — P25

The open responses support a narrower claim than “reference cases were useful.”
They functioned as **conditional and contested evidence**: useful when a
participant could establish clinically relevant similarity, but potentially
distracting when the relationship between cases was unclear or unconvincing.
P24 additionally showed that the reference case was only one element in a
larger evidential configuration that also included visible ECG features and the
consistency score.

**Boundary:** the qualitative responses alone cannot establish whether
reference cases generally helped or burdened users.

**Design implication:** the interface should align corresponding leads,
annotate discriminating features, and state why a reference case was retrieved,
so that similarity can become actionable evidence.

### Theme 5 — Usability, learning curve, and perceived training potential

**Coverage:** `r theme_summary$n_participants[theme_summary$theme_id == "T5"]` of
`r QUALITATIVE_N` qualitative contributors.

Five participants described the interface as easy or easier after
familiarization. Four named concrete friction: explanations were long, the
system was slow, x-axis changes were initially confusing, the first use
required familiarization, or the reference display was distracting. Four
responses framed the tool as supporting learning, revision, or early-career
use.

> “Very easy to use, very long explanation” — P03

> “it's just a bit slow and the switches between the x-axis size were a bit confusing in the beginning” — P09

> “it was a bit confusing to start, however after getting used to it, i found it easier to use.” — P19

> “instructive and helped revise my branch blocks” — P23

**Boundary:** these data support perceived educational potential, not improved
learning, retention, or transfer.

**Design implication:** use progressive disclosure, keep axes stable, shorten
explanations, and distinguish a training mode from a time-sensitive
point-of-care mode.

## Integrated qualitative account

The strongest cross-theme story is the contrast between answer-like compression
and comparison-based verification. Direct predictions were described as
narrowing the task and directing attention. Reference examples were evaluated
more conditionally: they became meaningful when they corroborated another
judgment, but could distract or reduce trust when perceived as dissimilar.
Trust connected the two mechanisms. Participants neither uniformly accepted nor
uniformly rejected AI; they used their own interpretations, observed model
behavior, manual checks, interface cues, and beliefs about model development to
decide how much weight to give the support.

Four higher-level propositions follow from the thematic and configurational
analyses:

1. **Diagnostic compression and trust are related but distinct.** Prediction
   could narrow the task even when participants continued to verify its output.
2. **Calibration practices and reliance vulnerabilities coexist.** The largest
   theme overlap is between trust calibration and guidance-versus-substitution,
   indicating that checking behavior does not eliminate influence.
3. **A retrieved case becomes evidence only through interpreted relevance.**
   Participants described both corroboration and distraction; a raw comparison
   was not self-explanatory support.
4. **Interaction quality and epistemic trust are separable.** A system could
   become easier to operate while still attracting skepticism about what its
   model can represent.

::: {.callout-important title="Paper-ready qualitative result"}
Participants described direct predictions as diagnostic-navigation aids that
narrowed possible differentials, supplied an initial direction, and focused
attention on particular ECG features. Trust was nevertheless conditional:
respondents cited agreement with their own interpretation, observed accuracy,
manual verification, displayed prediction stability or model agreement,
training-data provenance, and prior experience with clinical AI. These accounts
also revealed a tension between verification and cognitive substitution. Some
participants emphasized double-checking, while others explicitly linked limited
ECG knowledge to increased reliance or reported that AI influenced diagnostic
thresholds or tempted them to omit a step-by-step analysis. Reference cases
were described as conditional and contested evidence: similarity could support
trust, while perceived mismatch or distraction could undermine it. Finally,
participants
identified specific interface friction, including lengthy explanations,
slowness, changing axes, and an initial learning curve.
:::

## Claims and evidential boundaries {.unnumbered}

| Supported by the qualitative data | Not supported by the qualitative data |
|---|---|
| Predictions were perceived as narrowing diagnostic search and directing attention. | The study demonstrated automation bias or behavioral overreliance. |
| Trust was described as calibrated through several clinical, experiential, and interface cues. | The tool caused participants to abandon step-by-step ECG analysis. |
| Some responses reveal a tension between active verification, bounded reliance, and possible cognitive substitution. | Reports of shortcut risk establish that participants actually skipped required reasoning steps. |
| Reference cases sometimes corroborated trust, while perceived mismatch or distraction could undermine it. | Reference cases are generally ineffective, burdensome, or helpful. |
| Concrete usability friction and perceived training potential were present. | The tool improved learning, retention, or transfer. |
| Themes organize this small prompted corpus. | The themes reached saturation or represent population prevalence. |

## Export reproducibility outputs {.unnumbered}

The following block writes analysis outputs to
`qualitative_analysis_outputs`. It does not alter the original survey.

```{r}
#| label: export-reproducibility-outputs

participant_code_map <- aggregate(
  code ~ participant + participant_label,
  data = unique(coding[, c("participant", "participant_label", "code")]),
  FUN = collapse_unique
)

participant_theme_map <- aggregate(
  theme ~ participant + participant_label,
  data = unique(coding_theme[, c("participant", "participant_label", "theme")]),
  FUN = collapse_unique
)

participant_summary <- aggregate(
  response_available ~ participant + participant_label,
  data = responses_long_all,
  FUN = sum
)
names(participant_summary)[names(participant_summary) == "response_available"] <-
  "available_responses"
participant_summary$included_in_qualitative_analysis <-
  participant_summary$participant %in% QUALITATIVE_PARTICIPANTS

participant_summary <- merge(
  participant_summary,
  participant_code_map,
  by = c("participant", "participant_label"),
  all.x = TRUE,
  sort = TRUE
)
participant_summary <- merge(
  participant_summary,
  participant_theme_map,
  by = c("participant", "participant_label"),
  all.x = TRUE,
  sort = TRUE
)
participant_summary$code[is.na(participant_summary$code)] <- ""
participant_summary$theme[is.na(participant_summary$theme)] <- ""

missing_responses <- responses_long_all[
  !responses_long_all$response_available,
  ,
  drop = FALSE
]

participant_theme_matrix_export <- data.frame(
  participant_label = rownames(participant_theme_matrix),
  participant_theme_matrix,
  check.names = FALSE,
  row.names = NULL
)
theme_intersections_export <- data.frame(
  theme_id = rownames(theme_intersections),
  theme_intersections,
  check.names = FALSE,
  row.names = NULL
)
theme_jaccard_export <- data.frame(
  theme_id = rownames(theme_jaccard),
  round(theme_jaccard, 4),
  check.names = FALSE,
  row.names = NULL
)

write_utf8_csv(
  responses_long_all,
  file.path(output_dir, "qualitative_responses_long_all.csv")
)
write_utf8_csv(
  responses_long,
  file.path(output_dir, "qualitative_responses_long.csv")
)
write_utf8_csv(
  missing_responses,
  file.path(output_dir, "qualitative_missing_responses.csv")
)
write_utf8_csv(
  codebook,
  file.path(output_dir, "qualitative_codebook.csv")
)
write_utf8_csv(
  coding,
  file.path(output_dir, "qualitative_coding_assignments.csv")
)
write_utf8_csv(
  coding_theme,
  file.path(output_dir, "qualitative_coding_with_themes.csv")
)
write_utf8_csv(
  code_summary,
  file.path(output_dir, "qualitative_code_summary.csv")
)
write_utf8_csv(
  theme_definitions,
  file.path(output_dir, "qualitative_theme_definitions.csv")
)
write_utf8_csv(
  theme_map,
  file.path(output_dir, "qualitative_theme_map.csv")
)
write_utf8_csv(
  theme_summary,
  file.path(output_dir, "qualitative_theme_summary.csv")
)
write_utf8_csv(
  quotes,
  file.path(output_dir, "qualitative_exemplar_quotes.csv")
)
write_utf8_csv(
  uncoded_responses,
  file.path(output_dir, "qualitative_uncoded_responses.csv")
)
write_utf8_csv(
  participant_summary,
  file.path(output_dir, "qualitative_participant_summary.csv")
)
write_utf8_csv(
  code_question_summary,
  file.path(output_dir, "qualitative_code_by_question.csv")
)
write_utf8_csv(
  participant_theme_configuration,
  file.path(output_dir, "qualitative_participant_theme_configuration.csv")
)
write_utf8_csv(
  participant_theme_matrix_export,
  file.path(output_dir, "qualitative_participant_theme_matrix.csv")
)
write_utf8_csv(
  theme_intersections_export,
  file.path(output_dir, "qualitative_theme_intersections.csv")
)
write_utf8_csv(
  theme_jaccard_export,
  file.path(output_dir, "qualitative_theme_jaccard.csv")
)
write_utf8_csv(
  theme_overlap_pairs,
  file.path(output_dir, "qualitative_theme_overlap_pairs.csv")
)
write_utf8_csv(
  evidence_source_summary,
  file.path(output_dir, "qualitative_trust_evidence_sources.csv")
)
write_utf8_csv(
  contrast_cases,
  file.path(output_dir, "qualitative_contrasting_cases.csv")
)

exported_files <- c(
  "qualitative_responses_long_all.csv",
  "qualitative_responses_long.csv",
  "qualitative_missing_responses.csv",
  "qualitative_codebook.csv",
  "qualitative_coding_assignments.csv",
  "qualitative_coding_with_themes.csv",
  "qualitative_code_summary.csv",
  "qualitative_theme_definitions.csv",
  "qualitative_theme_map.csv",
  "qualitative_theme_summary.csv",
  "qualitative_exemplar_quotes.csv",
  "qualitative_uncoded_responses.csv",
  "qualitative_participant_summary.csv",
  "qualitative_code_by_question.csv",
  "qualitative_participant_theme_configuration.csv",
  "qualitative_participant_theme_matrix.csv",
  "qualitative_theme_intersections.csv",
  "qualitative_theme_jaccard.csv",
  "qualitative_theme_overlap_pairs.csv",
  "qualitative_trust_evidence_sources.csv",
  "qualitative_contrasting_cases.csv"
)

data.frame(exported_file = exported_files, stringsAsFactors = FALSE) |>
  knitr::kable(caption = "Reproducibility outputs created by this notebook")
```
