---
title: "Decision Support Tool - Time Spent"
---

```{r}
#| message: false
#| warning: false
#| code-fold: true
#| code-summary: "Libraries, helper functions"
library(lme4)
library(lmerTest)
library(emmeans)
library(DHARMa)
library(ggplot2)


# A function to put lmer results in a markdown table
lmer_markdown <- function(model, caption = "") {
    s <- summary(model)
    coefs <- as.data.frame(s$coefficients)

    CI_lower <- coefs$Estimate - 1.96 * coefs$`Std. Error`
    CI_upper <- coefs$Estimate + 1.96 * coefs$`Std. Error`
    ratio <- exp(coefs$Estimate)

    res_df <- data.frame(
        Term = rownames(coefs),
        Estimate = sprintf("%.4f", coefs$Estimate),
        `Std. Error` = sprintf("%.4f", coefs$`Std. Error`),
        `95% CI` = sprintf("[%.4f, %.4f]", CI_lower, CI_upper),
        `Ratio (exp(b))` = sprintf("%.3f", ratio),
        `t value` = sprintf("%.3f", coefs$`t value`),
        check.names = FALSE
    )

    if ("Pr(>|t|)" %in% colnames(coefs)) {
        res_df$`p value` <- ifelse(coefs$`Pr(>|t|)` < 0.001, "< 0.001", sprintf("%.4f", coefs$`Pr(>|t|)`))
    } else {
        p_vals <- 2 * (1 - pnorm(abs(coefs$`t value`)))
        res_df$`p value` <- ifelse(p_vals < 0.001, "< 0.001", sprintf("%.4f", p_vals))
    }

    knitr::kable(res_df, caption = caption, align = "lrrrrrr")
}
```

```{r}
raw_data <- read.csv("patient_timesums.csv")

# Ensure response is numeric (seconds spent)
raw_data$seconds <- as.numeric(raw_data$seconds)

# Filter out extreme outliers (> 30 minutes)
clean_data <- raw_data[raw_data$seconds <= 1800, ]

# Ensure random grouping variables are factors
clean_data$participant <- factor(clean_data$participant)
clean_data$patient <- factor(clean_data$patient)
clean_data$arm_name <- factor(clean_data$arm_name)
clean_data$trial_order <- clean_data$patient_index

# 2x2 Factorial coding
clean_data$pred_support <- factor(ifelse(clean_data$arm %in% c(1, 3), "Yes", "No"), levels = c("No", "Yes"))
clean_data$case_support <- factor(ifelse(clean_data$arm %in% c(2, 3), "Yes", "No"), levels = c("No", "Yes"))

if (knitr::is_html_output()) {
    rmarkdown::paged_table(clean_data)
}
```

## Boxlot

Distribution of time spent (in seconds) across study arms:

```{r}
#| fig-width: 8
#| fig-height: 4.5
ggplot(clean_data, aes(x = seconds, y = reorder(arm_name, seconds, FUN = median), fill = arm_name)) +
    geom_boxplot(alpha = 0.6, outlier.shape = NA) +
    geom_jitter(height = 0.15, alpha = 0.35, size = 1.8) +
    scale_x_log10(breaks = c(10, 30, 60, 120, 300, 600, 1200)) +
    theme_minimal(base_size = 12) +
    labs(
        title = "Distribution of Time Spent per Study Arm",
        x = "Time spent per patient in Seconds (log scale)",
        y = "Study Arm"
    ) +
    theme(
        legend.position = "none",
        panel.grid.minor = element_blank()
    )
```

## Linear Mixed Model on Time Spent

Fits a a linear mixed model with crossed random intercepts for participant and patient.
Trial order and experience is added as a covariate.


```{r}
model_seconds <- lmer(
    log(seconds) ~ pred_support * case_support + trial_order + experience_level + (1 | participant) + (1 | patient),
    data = clean_data,
)

lmer_markdown(model_seconds, caption = "Model 1: Linear Mixed Model on Log Time Spent (seconds)")
```
**Trial order:** Appears to have a negative significant effect.
**Experience**: Not significant

## Estimated Marginal Means & Pairwise Contrasts

Estimated mean task completion time (seconds) per study arm back-transformed from log scale:

```{r}
emm_arms <- emmeans(model_seconds, ~ pred_support * case_support, type = "response")
emm_arms

pairs(emm_arms)
```

## Is the model fitted correctly?

It appears so.

```{r}
sim_res <- simulateResiduals(model_seconds)
plot(sim_res)
```